---
title: "Bioinformatics in R. Lab 2: DSeq2"
author: "J. Cesar Ignacio Espinoza - Cesar   "
date: "Week 02: March 25 and 27, 2024"
output: 
  html_document: 
    highlight: espresso
    theme: cerulean
---

```{r setup}
    ### Edit this line so your notebook renders properly
    knitr::opts_knit$set(root.dir = normalizePath("~/Bioinformatics")) 
```

## Differential Gene Expression From Counts Table to DeSeq2
Begin by downloading the data from (Same link as in today's canvas page), download to your working directory.

#Install the required librries and load the data.

Look at the code block below, it will install DESEQ2
```{r}
#if (!require("BiocManager", quietly = TRUE))
    #install.packages("BiocManager")

#BiocManager::install("DESeq2")
```

We now load the libraries
```{r message=FALSE, warning=FALSE, paged.print=FALSE, results="hide"}
# We first need to import the important library for today's class, dplyr
library(dplyr)
library(readr)
library(DESeq2)
library(ggplot2)

```
You can import data into DESeq2 from multiple sources, specifically:

- From transcript abundance files and tximport
- From a count matrix
- From htseq-count files
- From a SummarizedExperiment object

We will be uploading from a counts table:
```{r}
### Run this chunk to import the counts table and metadata into your evironment.
counts <- read.csv('airway_scaledcounts.csv', row.names = 1, header = TRUE)
metadata <- read.csv('airway_metadata.csv', row.names = 1, header = TRUE)

View(metadata)
View(counts)
```

**Side Quest** From Monday Excercise, open a text exitor or even excel! Anc create a metadata file from the small dataset that you ran monday (solution posted on canvas, feel free to run it if you need to do that again!)

**Main Story** Explore your data, both the counts table and the metadata.

```{r}
### Create a new data object called it sample_info, use the same information as metadata but remove the GEO_ID Column, and change the name from dex to dexamethasone in the header, show your code here

sample_info <- metadata %>%
  select(-geo_id) %>%
  dplyr::rename(dexamethasone = dex)
sample_info

```

You have all you need to run DESeq2, a counts table and a metadata. 

**SideQuest** Look into the console and type DESeqDataSet and see what options you get, create a code chunk here and write the different functions as comments with a couple of words describing the kind of input they take!


```{r}
dds <- DESeqDataSetFromMatrix(countData = counts, 
                              colData = sample_info,
                              design = ~ dexamethasone)
```

And after that you have your dds object!!

Take a look at it!

```{r}

View(dds)
# You can recover the original counts matrix with the counts function!
#counts(dds)
```

### It is a good idea to remove cells with low counts, why?
HINT: Think about it from the statistics point of view

Thinking about outliers, could be error, but also need significance for our values

#### Students that took R already, use base R to subset the dds object based on the sum across a row, 
say I just want to keep genes with > than 10 reads across al samples!?!

#### Students that haven;t take R think about a pseudocode way to do it!
```{r}
#count(dds) ##error in i 
subset_dds <- dds[rowSums(counts(dds)) > 10]
subset_dds
```

### How many genes are retained? 
i.e. how many passed the filtering step?

**Answer Here**:

Set a factor level:
#### Important, DESeq2 will need a reference (think about it as the baseline, wildtype, untreated, control, etc...)
It will normally take the first one alfabetical, but to make sure you can use the relevel function! Look at this example, edit
according to what you are working with
```{r}
#relevel(data$drug, ref = 'control')
relevel(dds$dexamethasone, ref = 'control')
```

Now actually run DESEQ

```{r}
dds <- DESeq(subset_dds)
```

Look at the output as it runs! This is actually running on your computer, so depending on what you have it might take 10 seconds or a couple of minutes

### Let's Start looking at the data:
run counts() again on the object created from the main DESEq function, create two code chunks, run the counts() method twice but add normalzied=T to one of them, look at the diffrences, which ones are scaled up? which ones are scaled down?
```{r}
### counts()
### counts()
```

We are now ready to look at our results! We apply the results function, print the help of the results function and select an alpha of 0.05
```{r}
res <- results(dds, alpha = 0.05)
res
```
### And that's it! you have a set of genes and how differentially expressed these genes are!
Use the summary() function to look at your data 
```{r}
summary(res)

```

How many genes are upregulaed?? How many are downregulated??

Use the order() fucntion to sor the res object by adjusted p values

# Intro to Viz:

M stands for "minus" or "log ratio": On the horizontal axis (x-axis), you'll see the expression levels of genes in one condition (like healthy cells), subtracted from the expression levels in another condition (like diseased cells). This gives you a measure of how much the gene expression changes between the two conditions.

A stands for "average" or "average abundance": On the vertical axis (y-axis), you'll see the average expression levels of genes from both conditions. This helps to see if there's any bias or systematic difference between the two conditions.

```{r}
plotMA(dds)
```

### This is it guys! Save your notebook and submit it for credit for today's activity! Monday we will be looking at more complex visuzaliations with DESeq outputs!

```{r}
BiocManager::install("gprofiler")
BiocManager::install("EnhancedVolcano")
```

```{r}
library(EnhancedVolcano)
sigs.df = as.data.frame(res)
EnhancedVolcano(sigs.df,x = 'log2FoldChange', y = 'padj',  lab= rownames(sigs.df))
```
```{r}
## from manual 
 EnhancedVolcano(res,
    lab = rownames(res),
    x = 'log2FoldChange',
    y = 'pvalue',
    title = 'N061011 versus N61311',
    pCutoff = 10e-32,
    FCcutoff = 0.5,
    pointSize = 3.0,
    labSize = 6.0)
```

