---
title: "Final Bioinformatic Project"
author: "Ochoa"
date: "2024-04-20"
output: html_document
---

```{r setup, include=FALSE}
     knitr::opts_knit$set(root.dir = normalizePath("~/Bioinformatics")) 
```

```{r Prepping Libraries }
#BiocManager::install("org.Mm.eg.db") 
library(dplyr)
library(readr)
library(DESeq2)
library(ggplot2)
library(EnhancedVolcano)
library("org.Mm.eg.db")
library(ComplexHeatmap)
library(clusterProfiler)
library(gprofiler2) 
```
#### Original Data in a BASH, but utilizing JN counts table and metadata 
```{r counts table}
PRcounts = read.table('JNcounts.txt', row.names = 1, header = TRUE)
PRmetadata = read.csv('JNmetadata.csv', row.names = 1, header = TRUE)
```
```{r adjusting data frame, saved under updated name }
PRcounts = PRcounts %>%
  select(-Chr, -Start, -End, -Strand, -Length) 
```

```{r DESEQ Dataframe }
dds = DESeqDataSetFromMatrix(countData = PRcounts, 
                              colData = PRmetadata,
                              design = ~ Age)
## Subset deseq object
subset =  dds[(rowSums(counts(dds)) >= 10),]

### Base is for preferred treatment, setting Young age as 'ref/control'
relevel(dds$Age, ref='Young')

### Run deseq2
DeSeqProd = DESeq(subset)
#### Save the results to a new object
StatusRes = results(DeSeqProd, alpha = 0.05)
```
```{r Initial Visualization }
StatusRes
```

```{r Gene Symbol now with Gene Names }
SigDF = as.data.frame(StatusRes)
SigDF$symbol = mapIds(org.Mm.eg.db, keys= rownames(SigDF), keytype = 'ENSEMBL', colum = "SYMBOL")
```
#### For the following section, I needed to use a really high cut off value because there were no genes that made the adjusted p value for the observations we have. I did this so that we could see the value of the heat map, and I will attempt the GO Enrichment to see what happens 
```{r Heat Map}
Diff_df = as.data.frame(SigDF)
Diff_df = Diff_df %>%
  filter(padj < 0.5)
#, #log2FoldChange > 3.5, #baseMean > 100)

pmat = counts(DeSeqProd, normalized = T)[rownames(Diff_df),]
colnames(pmat) = c('MidAge PBS', 'MidAge SARS-CoV-02', 'Young PBS', 'Young SARS-CoV-02')
pmat.z = t(apply(pmat,1, scale))
colnames (pmat.z) = colnames(pmat)
Heatmap(pmat.z, cluster_rows= T, cluster_columns= T, name = "Z-score", row_labels = Diff_df[rownames(pmat.z),]$symbol, column_names_gp = gpar(fontsize = 10), row_names_gp = gpar(fontsize = 5), row_names_rot = 45, column_names_rot = 25)
```
####So Go Enrichment did not work... I suppose I can try again with sickness instead.
```{r}
Diff_df = as.data.frame(SigDF)
Diff_df = Diff_df %>%
  filter(padj < 0.05)

proj_tested_genes = rownames(Diff_df)
GO_proj_results = enrichGO(gene = proj_tested_genes, OrgDb = "org.Mm.eg.db", keyType = 'ENSEMBL', ont = "BP")
fit = plot(barplot(GO_proj_results, showCategory=12, font.size = 8))
```

```{r DeSeq Round 2}
dds = DESeqDataSetFromMatrix(countData = PRcounts, 
                              colData = PRmetadata,
                              design = ~ Status)

subset =  dds[(rowSums(counts(dds)) >= 10),]

relevel(dds$Status, ref='Control')

DeSeqProd = DESeq(subset)

StatusRes = results(DeSeqProd, alpha = 0.05)

StatusRes

SigDF = as.data.frame(StatusRes)
SigDF$symbol = mapIds(org.Mm.eg.db, keys= rownames(SigDF), keytype = 'ENSEMBL', colum = "SYMBOL")
```

```{r Reattempting Heat Map }
Diff_df = as.data.frame(SigDF)
Diff_df = Diff_df %>%
  filter(padj < 0.05, log2FoldChange > 3.5, baseMean > 100)

pmat = counts(DeSeqProd, normalized = T)[rownames(Diff_df),]
colnames(pmat) = c('MidAge PBS', 'MidAge SARS-CoV-02', 'Young PBS', 'Young SARS-CoV-02')
pmat.z = t(apply(pmat,1, scale))
colnames (pmat.z) = colnames(pmat)
Heatmap(pmat.z, cluster_rows= T, cluster_columns= T, name = "Z-score", row_labels = Diff_df[rownames(pmat.z),]$symbol, column_names_gp = gpar(fontsize = 8), row_names_gp = gpar(fontsize = 8), row_names_rot = 45, column_names_rot = 45)
```

```{r GO ENRICHMENT}
Diff_df = as.data.frame(SigDF)
Diff_df = Diff_df %>%
  filter(padj < 0.05)

proj_tested_genes = rownames(Diff_df)
GO_proj_results = enrichGO(gene = proj_tested_genes, OrgDb = "org.Mm.eg.db", keyType = 'ENSEMBL', ont = "BP")
fit = plot(barplot(GO_proj_results, showCategory=10, font.size = 8))
```

```{r Plot GSEA}
go_projupreg <- gost(proj_tested_genes, organism = 'mmusculus', significant = T)

res2 <- projstatus_res[order(-projstatus_res$stat),]
proj_gene_list <- res2$stat
names(proj_gene_list) <- rownames(res2)
gsea_plots <- gseGO(proj_gene_list, ont = "BP", keyType = "ENSEMBL", OrgDb ="org.Mm.eg.db", eps = 1e-100 )
gseaplot(gsea_plots, geneSetID = "GO:0046324" )

##changed boundary for calculating p value 
```

```{r}
EnhancedVolcano(SigDF, x='log2FoldChange', 
                y= 'padj', 
                lab = SigDF$symbol,
                title = 'Fold Change and Significance of Expressed Genes',
                pCutoff = 10e-5,
                FCcutoff = 2,
                pointSize = 1.0,
                labSize = 2.0,
                colAlpha = 1)
```

