---
title: "Bioinformatics in R. Lab 2: Data Visualization and GO analysis"
author: "J. Cesar Ignacio Espinoza - Cesar   "
date: "Week 03: April 1st and 3rd 2024"
output: 
  html_document: 
    highlight: espresso
    theme: cerulean
---

```{r setup}
    ### Edit this line so your notebook renders properly
    knitr::opts_knit$set(root.dir = normalizePath("~/Bioinformatics")) 
```

## Differential Gene Expression From Counts Table to DeSeq2
Begin by downloading the data from (Same link as in today's canvas page), download to your working directory.

#Install the required librries and load the data.

You will have to install enhanced volcano, if this gives issues like last time, we can try a different software.
```{r}
#BiocManager::install("EnhancedVolcano")
```

We now load the libraries
```{r message=FALSE, warning=FALSE, paged.print=FALSE, results="hide"}
# We first need to import the important libnrary for today's class, dplyr
library(dplyr)
library(readr)
library(DESeq2)
library(ggplot2)
library(EnhancedVolcano)

```

Here we will run the entire dseq program to find significantly expressed genes

```{r}
### Run this chunk to import the counts table and metadata into your evironment.
counts <- read.csv('airway_scaledcounts.csv', row.names = 1, header = TRUE)
metadata <- read.csv('airway_metadata.csv', row.names = 1, header = TRUE)
### Modify the metadata table
sample_info <- metadata %>%
  dplyr::select(-geo_id) %>%
  dplyr::rename(dexamethasone = dex)
sample_info

#creat a deseqobject

dds <- DESeqDataSetFromMatrix(countData = counts, 
                              colData = sample_info,
                              design = ~ dexamethasone)
## Subset your ddseq object
subset <-  dds[(rowSums(counts(dds)) >= 10),]

### fix the base to a preferred treatment
relevel(dds$dexamethasone, ref='control')

### Run deseq2

deseq_ob <- DESeq(subset)

#### Save the results to a new object
res <- results(deseq_ob, alpha = 0.05)
```

### New packages: 
This package will be able to convert between ENSEMBL ID and symbol (gene names) 

```{r}
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
BiocManager::install("org.Hs.eg.db")
```

```{r}
library("org.Hs.eg.db")
sigs.df <-  as.data.frame(res)
sigs.df$symbol <- mapIds(org.Hs.eg.db, keys= rownames(sigs.df), keytype = 'ENSEMBL', colum = "SYMBOL")
```

```{r}
sigs.df
```
### Make your first volcano plot!
```{r}
EnhancedVolcano(sigs.df, x='log2FoldChange', y = 'padj', lab = sigs.df$symbol)

```
Look at the manual (vignette for enhanced volcano) https://bioconductor.org/packages/devel/bioc/vignettes/EnhancedVolcano/inst/doc/EnhancedVolcano.html
Create a new code chunk and plot a better version of this volcano plot! (15 minutes)

```{r}
EnhancedVolcano(sigs.df, x= 'log2FoldChange', y ='padj', lab = sigs.df$symbol, boxedLabels = TRUE, col=c('green', 'blue', 'yellow', 'orange'), legendLabels=c('Not sig, no change in reg','Not sig, downreg/upreg','Significant, no change in reg', 'Significant, upreg/downreg'), legendPosition = 'left', legendLabSize = 15, legendIconSize = 5.0)
```

## Heatmaps!
```{R}
aBiocManager::install("ComplexHeatmap")
library(ComplexHeatmap)
```

```{R}
library(dplyr)
diff.df <- as.data.frame(sigs.df)
diff.df <- diff.df %>%
  filter(padj < 0.05)
```

```{R}
mat <- counts(deseq_ob, normalized = T)[rownames(diff.df),]
mat.z <- t(apply(mat,1, scale))
colnames(mat.z) <- colnames(mat)
Heatmap(mat.z, cluster_rows= T, cluster_columns= T, name = "Z-score", row_labels = diff.df[rownames(mat.z),]$symbol)
```
## Use dplyr to subset the table of significant values until you get a nice looking figure. 
(You cannot plot all genes at once so you have to subset in order to gain some insights from 
your data)
```{R}
diff.df <- as.data.frame(sigs.df)
diff.df <- diff.df %>%
  filter(padj < 0.05, log2FoldChange > 4, baseMean > 100)

mat <- counts(deseq_ob, normalized = T)[rownames(diff.df),]
mat.z <- t(apply(mat,1, scale))
colnames(mat.z) <- colnames(mat)
Heatmap(mat.z, cluster_rows= T, cluster_columns= T, name = "Z-score", row_labels = diff.df[rownames(mat.z),]$symbol)

```
## Gene Ontology Analysis
```{R}
BiocManager::install('clusterProfiler')
library(clusterProfiler)
```

```{R}
diff.df <- as.data.frame(sigs.df)
diff.df <- diff.df %>%
  filter(padj < 0.05, log2FoldChange > 2, baseMean > 100)

genes_to_test <- rownames(diff.df)
GO_results <- enrichGO(gene = genes_to_test, OrgDb = "org.Hs.eg.db", keyType = 'ENSEMBL', ont = "BP")
fit <- plot(barplot(GO_results, showCategory=10))
```

Another program is gost, we can try it below.
```{R}
library(gprofiler2) ##installed from packages side bar , not code 
go_upreg <- gost(genes_to_test, organism = 'hsapiens', significant = T)
gostplot(go_upreg, interactive = T)
```
It might be more informative to create plots for genes that are both upregulated and downregulated. Subset your data above and plot the GO enrichment for both upregulated and downreghulated.
```{R}

```
GSEA Plot, make one, Cesar will explain what it means 
```{R}
res <- res[order(-res$stat),]
gene_list <- res$stat
names(gene_list) <- rownames(res)
gse <- gseGO(gene_list, ont = "BP", keyType = "ENSEMBL", OrgDb ="org.Hs.eg.db", eps = 1e-300 )
gseaplot(gse, geneSetID = "GO:0046324" )

```